---
format: gfm
default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file and re-render. -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rollr2

<!-- badges: start -->
[![R-CMD-check](https://github.com/Felixmil/rollr2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Felixmil/rollr2/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Felixmil/rollr2/graph/badge.svg)](https://app.codecov.io/gh/Felixmil/rollr2)
<!-- badges: end -->

`rollr2` parses and rolls tabletop RPG dice notation such as `"2d20+2"`. From a single notation string it reports every individual die, the kept dice, and the adjusted total, and it can summarise the outcome distribution across many repeated rolls.

## Installation

Install the development version from GitHub:

```{r}
#| eval: false
# install.packages("pak")
pak::pak("Felixmil/rollr2")
```

## Roll once

`roll()` simulates a single roll and returns a `roll` object holding the individual dice and the total. The functions are stochastic, so fix a seed when you want reproducible output.

```{r}
library(rollr2)

set.seed(42)
roll("2d20+2")
```

A keep selector like `4d6h3` keeps only the highest (`h`) or lowest (`l`) dice; the printout then shows which dice were kept. The inverse drop spelling `4d6dl1` (drop the lowest of four d6, the conventional D&D ability-score roll) is the same thing said the other way round.

```{r}
set.seed(19)
roll("4d6h3")
```

A reroll marker like `2d6r2` rerolls once any die showing a value at or below the threshold (here 2), keeping the new value; `rr` instead rerolls until the die lands above the threshold. This is the D&D 5e Great Weapon Fighting style reroll.

```{r}
set.seed(7)
roll("2d6r2")
```

A notation can also sum several terms and constants, joined with `+` or `-`.

```{r}
set.seed(42)
roll("1d20+1d6+3")
```

A trailing comparator like `5d10>=8` counts how many dice meet the target instead of summing them, so the outcome is a success count rather than a total.

```{r}
set.seed(42)
roll("5d10>=8")
```

## Summarise a distribution

`roll_distribution()` rolls the whole notation many times and summarises the distribution of totals, printing a count per outcome and a text histogram.

```{r}
set.seed(42)
roll_distribution("2d6", n = 1000)
```

## Learn more

See `vignette("rollr2")` to get started, and the "Visualising roll distributions" article on the [package website](https://felixmil.github.io/rollr2/) for ggplot2 visualisations of roll outcomes.
