---
title: "Get started with rollr2"
vignette: >
  %\VignetteIndexEntry{Get started with rollr2}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
library(rollr2)
```

`rollr2` turns tabletop RPG dice notation into simulated rolls. This article walks through rolling once and reading the result, summarising the distribution of many rolls, and the full notation grammar the parser accepts.

Both `roll()` and `roll_distribution()` draw random numbers, so their output varies run to run. Fix a seed with `set.seed()` whenever you want reproducible results, as this article does throughout.

## Roll once

`roll()` takes a single notation string and returns a `roll` object. Printing it shows the notation, the individual dice, and the adjusted total.

```{r}
set.seed(1)
r <- roll("2d20+2")
r
```

A `roll` object is a list. The three fields you will use most are:

- `dice`: an integer vector of every die rolled.
- `kept`: the dice that contribute to the total (equal to `dice` when there is no keep selector).
- `total`: the sum of the kept dice plus the modifier.

```{r}
r$dice
r$kept
r$total
```

When the notation carries a keep selector, `kept` is a subset of `dice` and the printout gains a `Kept:` line. Here `4d6h3` rolls four six-sided dice and keeps the highest three.

```{r}
set.seed(1)
kept_roll <- roll("4d6h3")
kept_roll

kept_roll$dice
kept_roll$kept
kept_roll$total
```

## Summarise many rolls

`roll_distribution()` rolls the whole notation `n` times and summarises the distribution of totals. Its print method shows the possible total range, a count per observed outcome, and a text histogram.

```{r}
set.seed(1)
roll_distribution("2d6", n = 1000)
```

The returned `roll_distribution` object stores the observed totals in `counts`, a named integer vector whose names are the outcome totals (ascending, with zero-count outcomes dropped), plus the theoretical `range` of possible totals.

```{r}
d <- roll_distribution("2d6", n = 1000)
d$counts
d$range
```

## Notation grammar

The parser accepts the following forms. The leading `d` is case-insensitive and surrounding whitespace is tolerated.

| Form | Meaning | Example |
|------|---------|---------|
| `NdX` | Roll `N` dice with `X` sides | `3d6` |
| `dX` | Count omitted, defaults to one die | `d20` |
| `NdX+M` | Add a modifier `M` to the total | `2d20+2` |
| `NdX-M` | Subtract a modifier `M` from the total | `1d8-1` |
| `NdXhK` | Keep the highest `K` dice | `4d6h3` |
| `NdXlK` | Keep the lowest `K` dice | `3d6l2` |
| `NdXh` / `NdXl` | Keep count omitted, defaults to keeping one die | `2d20h` |

The keep selector follows the die size and comes before the modifier, so a full notation reads count, die size, optional keep selector, optional modifier, for example `4d6h3+1`.

### Defaults

- A missing die count means one die (`d6` is the same as `1d6`).
- A missing modifier means no adjustment (`M = 0`).
- A missing keep count means keep one die (`2d20h` keeps the single highest die).

### Enforced limits

The parser rejects notation outside these bounds with a clear error:

- The die count must be a positive integer.
- The die size must be an integer of at least 2.
- The keep count must be between 1 and the die count (keeping zero dice, or more dice than were rolled, is rejected).

Non-integer counts, die sizes, and keep counts (such as `2.5d6` or `2d6h1.5`) do not parse.
